---
output: github_document
---

<!-- README.md is generated from README.Rmd. Please edit that file -->

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.path = "man/figures/README-",
  out.width = "100%"
)
```

# parseLatex

<!-- badges: start -->
<!-- badges: end -->

The goal of `parseLatex` is to provide a parser for a subset of
LaTeX syntax that is more complete than the `tools::parseLatex()` 
parser.  

Perhaps some day it will handle all LaTeX inputs, but that's not
likely.  For now, I'm aiming to handle anything that `knitr::kable()` 
and the `kableExtra` functions will produce.

At the moment, it doesn't even do as much as `tools::parseLatex()`.

## Installation

You can install the development version of parseLatex from [GitHub](https://github.com/) with:

``` r
# install.packages("pak")
pak::pak("dmurdoch/parseLatex")
```

## Example

This is a basic example :

```{r example}
library(parseLatex)
library(kableExtra)
latex <- kbl(mtcars[1:2, 1:2], format = "latex")
cat(latex)
parsed <- parseLatex(latex)
parsed
getParseData(parsed)
```
