---
output: github_document
---

<!-- README.md is generated from README.Rmd. Please edit that file -->

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.path = "man/figures/README-",
  out.width = "100%"
)
```

# parseLatex

<!-- badges: start -->
<!-- badges: end -->

The goal of `parseLatex` is to provide a parser for a subset of
LaTeX syntax that is more complete than the `tools::parseLatex()` 
parser.  

Perhaps some day it will handle all LaTeX inputs, but that's not
likely.  For now, I'm aiming to handle anything that `knitr::kable()` 
and the `kableExtra` functions will produce, plus related code.

## Installation

You can install the development version of parseLatex from [GitHub](https://github.com/) with:

``` r
# install.packages("pak")
pak::pak("dmurdoch/parseLatex")
```

## Example

This is a basic example.

```{r example}
library(parseLatex)
library(kableExtra)
latex <- kbl(mtcars[1:2, 1:2], format = "latex")
cat(latex)
parsed <- parseLatex(latex)
# This is a blank followed by a table; drop the blank
table <- parsed[[find_env(parsed, "tabular")]]
# Get the alignment options from the content
brace_options(get_contents(table))
tableCell(table, 2,2) # The title counts!
tableCell(table, 2,2) <- "Changed!"
table
```

## To-do list

- Should handle UTF-8 inputs, not just ASCII.
- Needs a vignette to describe how to use it.
- Possibly add class for table objects, and implement
  indexing and subsetting on that class.
- See what `kableExtra` does with tables, and try to make that
easier.
- Allow catcodes to be specified before parsing. 
- Do something with math environments?
